---
title: "Non-linear relationships"
format: html
editor: source
---

The `01-plan/01-study-design/02-data-simulation.qmd` documents the original analysis plan. In addition, the `02-code/analysis-template.qmd` provides code for how we planned to analyse the data and test our hypotheses. More specifically, in these plans, we had assumed that, on a ln-scale, we could model the relationship between nitrogen ($N$) and total biomass ($B$). However, following data exploration `02-code/01-exploratory-data-analysis.qmd`, it was clear that even on a log-scale, the relationship was not adequately described by a linear model.

## Prepare environment

```{r}
# load the prep_env script
source(here::here("functions/prep_exp_env.R"))
```

## Testing for non-linearity

To show that the data for both experiment 1 and experiment 2 are better described by non-linear functions, we used cross-validation. Specifically, we fit basis-splines of increasing complexity for the log(N) variable and compared the results using leave-one-out cross-validation. When the fit was better for the non-linear models than for the linear model, we concluded that there was justification for using the non-linear models.

```{r}
# load the raw data
dat_raw <- readr::read_csv2(here::here("data-raw/all-raw-data-response-phase.csv"),
                            guess_max = 1000)
```

### Experiment 1

#### Erigeron_canadensis

```{r}
# set the species
i_species_sel <- "Erigeron_canadensis"

# exclude_high_N
exclude_high_N <- FALSE

# clean the data
dat_ex1_mod <- 
  clean_exp1(data = dat_raw,
             i_species = i_species_sel,
             exclude_high_N = exclude_high_N) |>
  dplyr::mutate(log_B = log(B),
                N = N - min(N),
                P = factor(P),
                M = factor(M))

# check the data
head(dat_ex1_mod)
```

Fit the different models and compare them using leave-one-out cross-validation. From this, it is clear that the linear model is the worst fit to the data and substantially worse than the other models.

```{r}
# run the cross-validation procedure
ec_mod <- compare_models_loocv(data = dat_ex1_mod, formulas = model_list_exp1) 

# extract the table
ec_mod <- ec_mod$LOO_est

# check the model
ec_mod$species <- i_species_sel

# reorder the variables
ec_mod <- dplyr::select(ec_mod, species, model:AIC)

# check the table
ec_mod
```


#### Phytolacca_americana

```{r}
# set the species
i_species_sel <- "Phytolacca_americana"

# exclude_high_N
exclude_high_N <- FALSE

# clean the data
dat_ex1_mod <- 
  clean_exp1(data = dat_raw,
             i_species = i_species_sel,
             exclude_high_N = exclude_high_N) |>
  dplyr::mutate(log_B = log(B),
                N = N - min(N),
                P = factor(P),
                M = factor(M))

# check the data
head(dat_ex1_mod)
```

Fit the different models and compare them using leave-one-out cross-validation. From this, it is clear that the linear model is the worst fit to the data and substantially worse than the other models.

```{r}
# run the cross-validation procedure
pa_mod <- compare_models_loocv(data = dat_ex1_mod, formulas = model_list_exp1) 

# extract the table
pa_mod <- pa_mod$LOO_est

# check the model
pa_mod$species <- i_species_sel

# reorder the variables
pa_mod <- dplyr::select(pa_mod, species, model:AIC)

# check the table
pa_mod
```


#### Solidago_gigantea

```{r}
# set the species
i_species_sel <- "Solidago_gigantea"

# exclude_high_N
exclude_high_N <- FALSE

# clean the data
dat_ex1_mod <- 
  clean_exp1(data = dat_raw,
             i_species = i_species_sel,
             exclude_high_N = exclude_high_N) |>
  dplyr::mutate(log_B = log(B),
                N = N - min(N),
                P = factor(P),
                M = factor(M))

# check the data
head(dat_ex1_mod)
```

Fit the different models and compare them using leave-one-out cross-validation. From this, it is clear that the linear model is the worst fit to the data and substantially worse than the other models.

```{r}
# run the cross-validation procedure
sg_mod <- compare_models_loocv(data = dat_ex1_mod, formulas = model_list_exp1) 

# extract the table
sg_mod <- sg_mod$LOO_est

# check the model
sg_mod$species <- i_species_sel

# reorder the variables
sg_mod <- dplyr::select(sg_mod, species, model:AIC)

# check the table
sg_mod
```


#### Sporobolus_indicus

```{r}
# set the species
i_species_sel <- "Sporobolus_indicus"

# exclude_high_N
exclude_high_N <- FALSE

# clean the data
dat_ex1_mod <- 
  clean_exp1(data = dat_raw,
             i_species = i_species_sel,
             exclude_high_N = exclude_high_N) |>
  dplyr::mutate(log_B = log(B),
                N = N - min(N),
                P = factor(P),
                M = factor(M))

# check the data
head(dat_ex1_mod)
```

Fit the different models and compare them using leave-one-out cross-validation. From this, it is clear that the linear model is the worst fit to the data and substantially worse than the other models.

```{r}
# run the cross-validation procedure
si_mod <- compare_models_loocv(data = dat_ex1_mod, formulas = model_list_exp1) 

# extract the table
si_mod <- si_mod$LOO_est

# check the model
si_mod$species <- i_species_sel

# reorder the variables
si_mod <- dplyr::select(si_mod, species, model:AIC)

# check the table
si_mod
```

## Export the model comparison

```{r}
# bind the different tables together
cv_table <- dplyr::bind_rows(ec_mod, pa_mod, sg_mod, si_mod)

# arrange by species and then fit
cv_table <- dplyr::arrange(cv_table, species, RMSE)
```

```{r}
# save the data
saveRDS(object = cv_table,
        file.path(figure_table_path, 
                  paste0("cv-tbl-exp1", figure_suff("", exclude_high_N), ".rds")))
```


#### Revised plan

As is clear from this analysis of the model fit that the linear model (**m1**) is a consistently worse fit to the data than the models that incorporate basis splines for the relationship between nitrogen (N) and biomass (B). This is true for a range of metrics calculated via leave-one-out cross validation.

Originally, we aimed to test whether there was a difference in how natives and invasives respond to nitrogen and microbes. We set-up the model in such a way that the $beta_7$ coefficient provided the critical test of this hypothesis. Specifically:

+ If $\beta_7 > 0$, the plant-soil feedback becomes more positive (or less negative) with increasing N in invasives compared to natives.  
+ If $\beta_7 < 0$, the plant-soil feedback becomes more negative with increasing N in invasives compared to natives.

However, given that the relationship between N and B is non-linear, we have revised this hypothesis. Specifically, rather than trying to test whether the plant-soil feedback becomes more positive (i.e. less negative) or more negative (i.e. less positive), we test whether there is a difference in how natives and invasives respond to nitrogen in the presence and absence of microbes.

To do this, we compare two models using a likelihood ratio test (LRT) based on the chi-squared distribution:

+ **(1)** The full model with main effects of N, M and P (i.e. native versus invasive), all pairwise interactions and the three-way interaction and where N is modeled using basis-splines (specific splines chosen based on model fit)

+ **(2)** A reduced model with main effects of N, M and P (i.e. native versus invasive), all pairwise interactions but without the three-way interaction and where N is modeled using basis-splines (specific splines chosen based on model fit)

If the full model has a significantly higher log-likelihood, we conclude that there is a difference in how natives and invasives respond to nitrogen in the presence and absence of microbes.



### Make the tables for Experiment 2...













