---
title: "Individual species responses"
format: html
editor: source

params:
  invasive_species: "Erigeron_canadensis"
  exclude_highest_N: FALSE
---

### Prepare the environment

```{r}
# load the prep_env script
source(here::here("functions/prep_exp_env.R"))
```

### Load the relevant data

We start by loading the raw data from the response phase.

```{r}
# load the raw data
dat_raw <- readr::read_csv2(here::here("data-raw/all-raw-data-response-phase.csv"), guess_max = 1000)
```

Set-up the the focal species if running interactively.

+ Erigeron_canadensis
+ Phytolacca_americana
+ Solidago_gigantea
+ Sporobolus_indicus

```{r}
# create the relevant parameters if params not present in environment
# set_default_params(list(
  # invasive_species = "Sporobolus_indicus",
  # exclude_highest_N = FALSE
# ))

# set the parameters
i_species_sel <- params$invasive_species
```

### Clean the data

Next, we run some basic data cleaning which involves subsetting the native species individually so we can examine how they respond to microbes along with recoding some of the variables.

```{r}
# extract the relevant data
dat <-
  dat_raw |>
  dplyr::filter(Setup_IAN == "N",
                Invasive_species %in% c("All", i_species_sel))

# calculate total weight
dat$total_weight <- 
  with(dat,
       Shoot_dry_weight + Root_dry_weight)
  
# reorder the columns
dat <-
  dat |>
  dplyr::select(Soil_conditioning, Setup_IAN, Nitrogen_level,
                Repetition_nr, Species, Shoot_dry_weight, Root_dry_weight,
                total_weight)

# rename the columns
names(dat) <- c("M", "P", "N", "rep", "species", "B_shoot", "B_root", "B")

# convert the treatment variables to the values to match the data simulation
dat <-
  dat |>
  dplyr::mutate(M = ifelse(M == "C", 0, 1),
                P = ifelse(P == "N", 0, 1),
                N = round(log(N), 2))

# sort out the variable types
dat <-
  dat |>
  dplyr::mutate(N = N - min(N),
                P = factor(P),
                M = factor(M),
                log_B = log(B + 1))
```


### Visualise the data

```{r}
ind_plot <- function(x_dat, x_dat_sum, microbes, y_lims = NA) {
  
  species_cols <- c(
    "Agrostis"   = "#6B8E23",  # muted olive green
    "Centaurea"  = "#9370DB",  # dusty violet
    "Plantago"   = "#8F9779",  # soft grey-green
    "Ranunculus" = "#D8B365"   # muted golden yellow
    )
  
  # extract y_lims
  y_lim_range <-
    if (all(is.na(y_lims))) {
      range(c(x_dat$log_B, x_dat_sum$log_B_m)) + c(-0.05, 0.05)
      } else {
        y_lims
        }
  
  # title colour
  title_colour <-
    if (microbes == "No microbes") {
      "gray60"
      } else {
        "forestgreen"
        }
  
  ggplot() +
    geom_rect(mapping = aes(xmin = 2.99, xmax = 3.40, ymin = -Inf, ymax = Inf),
              colour = NA, alpha = 0.15, fill = "darkred") +
    ggbeeswarm::geom_quasirandom(data = x_dat |> dplyr::filter(M == microbes),
                                mapping = aes(x = N, y = log_B, colour = species),
                                width = 0.05) +
    geom_line(data = x_dat_sum |> dplyr::filter(M == microbes),
              mapping = aes(x = N, y = log_B_m, colour = species),
              show.legend = FALSE) +
    scale_colour_manual(values = species_cols) +
    scale_x_continuous(breaks = round(unique(x_dat[["N"]]), 1)) +
    scale_y_continuous(limits = y_lim_range) +
    ylab("ln(1 + Biomass) (mg)") +
    xlab("ln(Nitrogen) (mg)") +
    ggtitle(paste0(microbes))+
    theme_meta() +
    theme(legend.title = element_blank(),
          plot.title = element_text(hjust = 0.5, size = 10))
  
}
```

Use the function to create the individual plot:

```{r}
# sort out the labels
dat_plot <- fix_labels(x = dat)

# summarise the data
dat_plot_sum <-
  dat_plot |>
  dplyr::group_by(M, P, N, species) |>
  dplyr::summarise(log_B_m = mean(log_B)) |>
  dplyr::ungroup()

# get the y_lims
y_lims <- range(dat_plot$log_B) + c(-0.05, 0.05)

# make the plots

# no microbes
p_nm <- ind_plot(x_dat = dat_plot, x_dat_sum = dat_plot_sum, 
                 microbes = "No microbes", y_lims = y_lims)

# microbes
p_m <- ind_plot(x_dat = dat_plot, x_dat_sum = dat_plot_sum, 
                microbes = "Microbes", y_lims = y_lims)

# bind the plots using patchwork
p_nm_m <-
  p_nm + p_m + 
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "a", title = gsub("_", " ", i_species_sel),
                  theme = theme(plot.title = element_text(size = 11, face = "italic")))

# check the plot
p_nm_m

# export the figure
saveRDS(p_nm_m,
        file = file.path(figure_table_path,
                         paste0("fig-ind-response-", figure_suff(i_species_sel, FALSE), ".rds")))
```










